### 2014-08-06 doc/thesis/cases.tex
Starting the rewriting of the Typed Lua chapter
Fixing _ENV to include setmetatable, type, and require
Trying another formatting style to the section about Typed Lua and sketching the file userdata

\begin{enumerate}
\item require (modname) :
\begin{verbatim}
(string) -> ({any:any})
\end{verbatim}
Category 1.
\\
The function \textbf{require} returns exactly the same value that is
returned by loading \textbf{modname}.
However, if the returned value is $\Nil$, this value is
converted to $\True$.
This function is a primitive in the type system.
\item package.config :
\begin{verbatim}
string
\end{verbatim}
Category 1.
\item package.cpath :
\begin{verbatim}
string
\end{verbatim}
Category 1.
\item package.loaded :
\begin{verbatim}
{string:{any:any}}
\end{verbatim}
Category 2?
\item package.loadlib (libname, funcname) :
\begin{verbatim}
(string, string) -> (true)?
\end{verbatim}
Category 1?
\item package.path :
\begin{verbatim}
string
\end{verbatim}
Category 1.
\item package.preload :
\begin{verbatim}
{string:{any:any}}
\end{verbatim}
Category 2?
\item package.searchers :
\begin{verbatim}
{(any*) -> (any*)}
\end{verbatim}
Category 2?
\item package.searchpath (name, path [, sep [, rep]]) :
\begin{verbatim}
(string, string, string?, string?) -> (string)?
\end{verbatim}
Category 1.
\\
I guess the proper type should be:
\begin{verbatim}
(string, string) |
(string, string, string) |
(string, string, string, string?) -> (string)?
\end{verbatim}
\end{enumerate}

@@

\begin{enumerate}
\item table.concat (list [, sep [, i [, j]]]) :
\begin{verbatim}
({string|number}, string?, number?, number?) -> (string)
\end{verbatim}
Category 1.
\\
I guess the proper type should be:
\begin{verbatim}
({string|number}) |
({string|number}, string) |
({string|number}, string, number?) -> (string)
\end{verbatim}
\item table.insert (list, [pos,] value) :
\begin{verbatim}
({any}, any) -> ()
\end{verbatim}
Category 2.
\\
This function does not accept/discard extra parameters.
Perhaps due to the fact that the parameter \textbf{pos} is optional.
If \textbf{value} is $\TopStar$, the function inserts \textbf{pos}
in the \textbf{list}, that is, the input type is
\texttt{(\{any\}, any) -> ()}.
If \textbf{value} is $\Nil$, the function does not insert \textbf{pos}
in the \textbf{list}, that is, the input type is
\texttt{(\{any\}, number, any) -> ()},
but the element is not inserted in the list because \textbf{value}
is actually $\Nil$.
I guess the proper type should be:
\begin{verbatim}
({any}, any) | ({any}, number, any) -> ()
\end{verbatim}
\item table.pack (···) :
\begin{verbatim}
(any*) -> ({number:any, "n":number})
\end{verbatim}
Category 2.
\\
Even though \textbf{table.pack} gets any number of parameters of
any type, the $\Nil$ type generates a hole in the returned table.
\item table.remove (list [, pos]) :
\begin{verbatim}
({any}, number?) -> (any)
\end{verbatim}
Category 2.
\\
The return type depends on the type of the element that is removed
from the list.
\item table.sort (list [, comp]) :
\begin{verbatim}
({any}, nil|(any, any) -> (boolean)) -> ()
\end{verbatim}
Category 2.
\\
Do we really need grouping?
\begin{verbatim}
({any}, [(any, any) -> (boolean)]?) -> ()
\end{verbatim}
Grouping would be useful for defining types like this:
\begin{verbatim}
([(number) -> (number)] | (string) -> (string)) -> ({string})
\end{verbatim}
But we do not have intersection types anyway.
\item table.unpack (list [, i [, j]]) :
\begin{verbatim}
({any}, number?, number?) -> (any*)
\end{verbatim}
Category 2.
\\
The return type depends on the type of the elements that are stored in
the list.
I guess the proper type should be:
\begin{verbatim}
({any}) | ({any}, number, number?) -> (any*)
\end{verbatim}
\end{enumerate}

@@

\begin{enumerate}
\item io.close ([file]) :
\begin{verbatim}
(any?) -> (boolean) | (nil, string) | (true?, string, number)
\end{verbatim}
Category 1?
\\
Returns \texttt{(true?, string, number)} when the file handle
was created with \textbf{io.popen}.
The parameter \textbf{file} is a \textit{userdata} and I need
to fix that on the type system.
\item io.flush () :
\begin{verbatim}
() -> (boolean)
\end{verbatim}
Category 1.
\item io.input ([file]) :
\begin{verbatim}
(string|any|nil) -> (any)
\end{verbatim}
Category 1?
Here the type \texttt{any} refers to a \textit{userdata} and
I need to fix that on the type system.
\item io.lines ([filename ···]) :
\begin{verbatim}
(string|any|nil) -> ((any*) -> (any*))
\end{verbatim}
Category 1?
\\
In fact, this function returns an iterator. Need to check the return type.
Here the type \texttt{any} refers to a \textit{userdata} and
I need to fix that on the type system.
\item io.open (filename [, mode]) :
\begin{verbatim}
(string,
"r"|"w"|"a"|"r+"|"w+"|"a+"|"rb"|"wb"|"ab"|"r+b"|"w+b"|"w+b"|nil) ->
(any) | (nil, string, number)
\end{verbatim}
Category 1?
\\
Here the type \texttt{any} refers to a \textit{userdata} and
I need to fix that on the type system.
\item io.output ([file]) :
\begin{verbatim}
(string|any|nil) -> (any)
\end{verbatim}
Category 1?
\\
Here the type \texttt{any} refers to a \textit{userdata} and
I need to fix that on the type system.
\item io.popen (prog [, mode]) :
\begin{verbatim}
(string, "r"|"w"|nil) -> (any) | (nil, string, number)
\end{verbatim}
Category 1?
\\
Here the type \texttt{any} refers to a \textit{userdata} and
I need to fix that on the type system.
\item io.read (···) :
\begin{verbatim}
("*n"|"*a"|"*l"|"*L"|number*) -> (string|number*)
\end{verbatim}
Category 1.
\item io.tmpfile () :
\begin{verbatim}
() -> (any)
\end{verbatim}
Category 1?
\\
Here the type \texttt{any} refers to a \textit{userdata} and
I need to fix that on the type system.
\item io.type (obj) :
\begin{verbatim}
(value) -> ("file"|"closed file"|nil)
\end{verbatim}
Category 1.
\item io.write (···) :
\begin{verbatim}
(string|number*) -> (any) | (nil, string, number)
\end{verbatim}
Category 1?
\\
Here the type \texttt{any} refers to a \textit{userdata} and
I need to fix that on the type system.
\item file:close () : ?
\item file:flush () : ?
\item file:lines (···) : ?
\item file:read (···) : ?
\item file:seek ([whence [, offset]]) : ?
\item file:setvbuf (mode [, size]) : ?
\item file:write (···) : ?
\end{enumerate}

>>>>

\begin{enumerate}
\item require (modname) :
\begin{verbatim}
(string) -> (any)
\end{verbatim}
Category 1.
\\
The function \textbf{require} returns exactly the same value that is
returned by loading \textbf{modname}.
However, if the returned value is $\Nil$, this value is
converted to $\True$.
This function is a primitive in the type system.
\item package.config :
\begin{verbatim}
string
\end{verbatim}
Category 1.
\item package.cpath :
\begin{verbatim}
string
\end{verbatim}
Category 1.
\item package.loaded :
\begin{verbatim}
{string:{any:any}}
\end{verbatim}
Category 2?
\item package.loadlib (libname, funcname) :
\begin{verbatim}
(string, string) -> (true)?
\end{verbatim}
Category 1?
\item package.path :
\begin{verbatim}
string
\end{verbatim}
Category 1.
\item package.preload :
\begin{verbatim}
{string:{any:any}}
\end{verbatim}
Category 2?
\item package.searchers :
\begin{verbatim}
{(any*) -> (any*)}
\end{verbatim}
Category 2?
\item package.searchpath (name, path [, sep [, rep]]) :
\begin{verbatim}
(string, string, string?, string?) -> (string)?
\end{verbatim}
Category 1.
\\
I guess the proper type should be:
\begin{verbatim}
(string, string) |
(string, string, string) |
(string, string, string, string?) -> (string)?
\end{verbatim}
\end{enumerate}

@@

\begin{enumerate}
\item table.concat (list [, sep [, i [, j]]]) :
\begin{verbatim}
({string|number}, string?, number?, number?) -> (string)
\end{verbatim}
Category 1.
\\
I guess the proper type should be:
\begin{verbatim}
({string|number}) |
({string|number}, string) |
({string|number}, string, number?) -> (string)
\end{verbatim}
\item table.insert (list, [pos,] value) :
\begin{verbatim}
({any}, any) -> ()
\end{verbatim}
Category 2.
\\
This function does not accept/discard extra parameters.
Perhaps due to the fact that the parameter \textbf{pos} is optional.
If \textbf{value} is $\Value{*}$, the function inserts \textbf{pos}
in the \textbf{list}, that is, the input type is
\texttt{(\{any\}, any) -> ()}.
If \textbf{value} is $\Nil$, the function does not insert \textbf{pos}
in the \textbf{list}, that is, the input type is
\texttt{(\{any\}, number, any) -> ()},
but the element is not inserted in the list because \textbf{value}
is actually $\Nil$.
I guess the proper type should be:
\begin{verbatim}
({any}, any) | ({any}, number, any) -> ()
\end{verbatim}
\item table.pack (···) :
\begin{verbatim}
(any*) -> ({number:any, "n":number})
\end{verbatim}
Category 2.
\\
Even though \textbf{table.pack} gets any number of parameters of
any type, the $\Nil$ type generates a hole in the returned table.
\item table.remove (list [, pos]) :
\begin{verbatim}
({any}, number?) -> (any)
\end{verbatim}
Category 2.
\\
The return type depends on the type of the element that is removed
from the list.
\item table.sort (list [, comp]) :
\begin{verbatim}
({any}, nil|(any, any) -> (boolean)) -> ()
\end{verbatim}
Category 2.
\\
Do we really need grouping?
\begin{verbatim}
({any}, [(any, any) -> (boolean)]?) -> ()
\end{verbatim}
Grouping would be useful for defining types like this:
\begin{verbatim}
([(number) -> (number)] | (string) -> (string)) -> ({string})
\end{verbatim}
But we do not have intersection types anyway.
\item table.unpack (list [, i [, j]]) :
\begin{verbatim}
({any}, number?, number?) -> (any*)
\end{verbatim}
Category 2.
\\
The return type depends on the type of the elements that are stored in
the list.
I guess the proper type should be:
\begin{verbatim}
({any}) | ({any}, number, number?) -> (any*)
\end{verbatim}
\end{enumerate}

@@

\begin{verbatim}
userdata file
  close : (self) -> (boolean) | (nil, string) | (true?, string, number),
  flush : (self) -> (boolean),
  lines : (self, string|number*) -> (() -> (any*)),
  read : (self, string|number*) -> (string*),
  seek : (self, string?, number?) -> (number?),
  setvbuf : (self, string, number?) -> (true) | (nil, string, number),
  write : (self, string|number*) -> (self?)
end
\end{verbatim}

\begin{enumerate}
\item io.close ([file]) :
\begin{verbatim}
(file?) -> (boolean) | (nil, string) | (true?, string, number)
\end{verbatim}
Category 1?
\\
Returns \texttt{(true?, string, number)} when the file handle
was created with \textbf{io.popen}.
\item io.flush () :
\begin{verbatim}
() -> (boolean)
\end{verbatim}
Category 1.
\item io.input ([file]) :
\begin{verbatim}
(string|file|nil) -> (file)
\end{verbatim}
Category 1?
\item io.lines ([filename ···]) :
\begin{verbatim}
(string|file|nil) -> ((any*) -> (any*))
\end{verbatim}
Category 1?
\\
In fact, this function returns an iterator. Need to check the return type.
\item io.open (filename [, mode]) :
\begin{verbatim}
(string,
"r"|"w"|"a"|"r+"|"w+"|"a+"|"rb"|"wb"|"ab"|"r+b"|"w+b"|"w+b"|nil) ->
(file) | (nil, string, number)
\end{verbatim}
Category 1?
\item io.output ([file]) :
\begin{verbatim}
(string|file|nil) -> (file)
\end{verbatim}
Category 1?
\item io.popen (prog [, mode]) :
\begin{verbatim}
(string, "r"|"w"|nil) -> (file) | (nil, string, number)
\end{verbatim}
Category 1?
\item io.read (···) :
\begin{verbatim}
("*n"|"*a"|"*l"|"*L"|number*) -> (string|number*)
\end{verbatim}
Category 1.
\item io.tmpfile () :
\begin{verbatim}
() -> (file)
\end{verbatim}
Category 1?
\item io.type (obj) :
\begin{verbatim}
(value) -> ("file"|"closed file"|nil)
\end{verbatim}
Category 1.
\item io.write (···) :
\begin{verbatim}
(string|number*) -> (file) | (nil, string, number)
\end{verbatim}
Category 1?
\end{enumerate}

### 2014-08-06 doc/thesis/typedlua.tex
Trying another formatting style to the section about Typed Lua and sketching the file userdata

\section{Type Language}
\label{sec:types}

@@

Figure \ref{fig:typelang} presents the abstract syntax of the
Typed Lua types.
We split Typed Lua types into two categories:
\emph{first-level types} and \emph{second-level types}.
We use first-level types to represent Lua values and
second-level types to type expression lists, multiple assignments,
and function applications.
We will explain all Typed Lua types in more detail along this section.

\section{Typed Lua through examples}
\label{sec:examples}

The main goal of Typed Lua is to allow programmers to combine static
and dynamic typing in Lua, so Typed Lua is an extension to Lua 5.2
syntax that introduces optional type annotations, and class-based
object-oriented programming through the definition of classes,
interfaces, and modules.
An important feature of Typed Lua is its backwards compatibility with
Lua, that is, unannotated Typed Lua code is valid Lua code.
For this reason, the syntactic extensions introduced by Typed Lua
include new keywords that are not reserved words.
Appendix \ref{app:syntax} presents the complete syntax of Typed Lua.

\section{The design of Typed Lua}
\label{sec:design}

\subsection{Subtyping and Consistent-Subtyping}

@@

The consistent-subtyping rules are exactly the same as the subtyping rules,
except those that handle the dynamic type.

@@

>>>>

\section{Typed Lua}
\label{sec:typedlua}

Typed Lua allows statically typed Lua code to coexist and interact
with dynamically typed Lua code.
The Typed Lua compiler checks for type errors, but alwayas removes
the annotations, generating Lua code that runs in unmodified Lua
implementations.
The programmer can enjoy some of the benefits of static types even
without converting existing Lua modules to Typed Lua:
a dynamically typed module can export a statically typed interface,
and statically typed users of the module will have their use of the
module checked by the compiler.

Unlike gradual typing, Typed Lua does not insert run-time checks
between dynamically typed and statically typed parts of the program.
Unlike some optional type systems, the statically typed subset of
Typed Lua aims to be sound, so a later version of the Typed Lua
compiler can switch to gradual typing instead of just optional typing.

In this section we cover the design of Typed Lua:
In Section \ref{sec:types} we present the types that are available
in Typed Lua along with the subtyping and consistent-subtyping
relations.
In Section \ref{sec:examples} we present the design of Typed Lua
through code examples.
In Section \ref{sec:rules} we present the design of Typed Lua
through typing rules.

\subsection{Type Language}
\label{sec:types}

@@

Figure \ref{fig:typelang} presents the abstract syntax of the
Typed Lua types.
We split Typed Lua types into two categories:
\emph{first-level types} and \emph{second-level types}.
We use first-level types to represent Lua values and
second-level types to type expression lists, multiple assignments,
and function applications.

Figure \ref{fig:subtyping} presents the subtyping relation among
Typed Lua types.
Any first-level type is a subtype of $\Value$.
The rest of the subtyping relationship is standard:
union types are supertypes of their respective literal types,
function types are related by contravariance on the input
and covariance on the output, table types have width subtyping,
with depth subtyping on $\Const$ fields, tuple and vararg types
are covariant.

@@

Figure \ref{fig:consistent_subtyping} presents the consistent-subtyping
relation that allows the dynamic type $\Any$ interoperate between
dynamically typed and statically typed code.
The dynamic type $\Any$ is a subtype of $\Value$, but neither a
supertype nor a subtype of any other type.
Our consistent-subtyping relationship follows the standards defined
by the gradual typing of objects \citep{siek2007objects,siek2013mutable}.
In practice, we can pass a value of the dynamic type anytime we want
a value of some other type, and can pass any value where where a
value of the dynamic type is expected, but these operations are tracked
by the type system, and the programmer can choose to be warned about them.

@@

\subsection{Typed Lua through examples}
\label{sec:examples}

The main goal of Typed Lua is to allow programmers to combine static
and dynamic typing in Lua, so Typed Lua is an extension to Lua 5.2
syntax that introduces optional type annotations, and class-based
object-oriented programming through the definition of classes,
interfaces, and modules.
An important feature of Typed Lua is its backwards compatibility with
Lua, that is, unannotated Typed Lua code is valid Lua code.
For this reason, the syntactic extensions introduced by Typed Lua
include new keywords that are not reserved words.
Appendix \ref{app:syntax} presents the complete syntax of Typed Lua,
which we cover in this section through code examples.

\subsection{The typing rules of Typed Lua}
\label{sec:rules}
